Depression and suicidality as costly signals of need

### 2021-03-30 Paper.rmd
Edited the intro

# Introduction
Since Durkheim’s pioneering effort to use quantitative social science to explain suicide, varied attempts by anthropologists, psychologists, and sociologists and substantial investment by funding agencies, rates of depression and suicide attempts have remained largely unchanged over the last 100 years (confirm and cite), and there is still substantial uncertainty about why both are as common as they are (cite). Due to their high costs, the vast majority of these approaches have been built on the assumption that they are the result of internal dysfunction (cite examples). However, a subset of anthropologists and evolutionary theorists have challenged this idea, suggesting that despite the subjective suffering they cause, they may function as a means to limit negative fitness outcomes in times of substantial adversity (Firth + others).

@@

## Connection to Adversity
Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where this is the case, the onset of depression tends to be rapid, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

This strong connection between negative life events and depression along with the need and distress that likely accompanies these events has led to the suggestion that this relationship is causal (Kendler et al., 1999). Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, which shows that the connection is unlikely to be driven solely by depressed individuals creating situations where negative events are likely to be common (Hammen, 2005; Kendler et al., 1999). In addition to this, twin studies have shown that one's history of negative events is still a strong predictor when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict (Kendler & Baker, 2007; Kendler et al., 1999).

@@

Suicide attempts also tend to co-occur with social adversity (Hagen et al., 2008 + cite and describe), with evidence of this existing cross-culturally. Greif over the loss of a loved one (Firth; Devries et al. 2011), bullying (Alavi et al., 2017; Butler et al., 2019; Ford et al., 2017; Wang et al., 2020), and instances of social transgression (Firth; Malinowski; Syme and Hagen 2018; find Western citation to confirm) have all been reported to lead to suicidal behavior in both industrialized (confirm the transgression part) and traditional societies. As with depression, assault is a powerful risk factor for suicidal behavior, something which has been found in multiple large-scale multi-country studies (Devries et al. 2011; Ellsberg et al. 2008). Both sexual and non-sexual assault were also associated with more self-harm and suicidal behaviors among both indigenous Sami and majority Norwegian adolescents, with those who attempted suicides reporting more than twice the number of negative life events than those who did not (Reigstad and Kvernmo 2017). The connection between social adversity and suicide is also seen in the Human Relations Area Files (HRAF), in which 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone (Syme et al., 2016).

@@

Beyond resulting in the cost differential that allows for costly signaling in the first place, the costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial. Importantly, there are likely few forms of adversity that do not come with conflict. For example, breakups in both romantic and cooperative relationships are often preceded by conflict (Amato & Hohmann-Marriott, 2007), and surely involve conflict when one side wants reconciliation and the other does not. Likewise, the death of a loved one may also be a source of conflict as the survivor may turn to others for more support who might not want to invest more in the survivor (Hagen, 2003). Regardless of the source of conflict, greater conflict should correspond with greater conflicts of interests, something which would favor greater investment in manipulating others through fake signals. In these instances, less costly signals like verbal requests, facial sadness, and crying are likely to be unreliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals in those experiencing need they cannot resolve on their own. Importantly, the amount of conflict

In addition to signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression is suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

@@

Beneficial social changes are also reported in traditional societies. Unlike Western countries, where suicide is often viewed as pathological (Hidaka, 2012), members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness (Shostak, 1981). For example, attempts have been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers (Gutierrez de Pineda & Muirden, 1948; Hilger, 1957; Karsten, 1935; Tessmann, 1930; Wilson, 1960), suggesting that help is at least occasionally provided or reliably expected. In support of this, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, even though suicide was rarely if ever the main topic (Syme et al., 2016).

>>>>

# Introduction
Since Durkheim’s pioneering effort to use quantitative social science to explain suicide (cite), anthropologists, psychologists, and sociologists have made varied attempts at explaining depression and suicide. However, despite this effort and substantial investment by funding agencies, rates of depression and suicide attempts have remained largely unchanged over the last 100 years (confirm and cite), and there is still substantial uncertainty about why both are as common as they are (cite). Likely due to their high costs and the suffering they cause, the vast majority of these approaches have been built on the assumption that they are the result of internal dysfunction (cite examples). However, a subset of anthropologists and evolutionary theorists have challenged this idea, suggesting that despite the subjective suffering they cause, they may function as a means to limit negative fitness outcomes in times of substantial adversity (Firth + others).

@@

## Connection to Adversity
Likely the most reliable finding when it comes to depression and suicidal behavior is that most cases are closely tied to major negative life events (Hammen, 2005 + others; citations for suicide). Compared to non-depressed individuals, those with depression tend to report at least twice as many instances of negative events in both community and clinical samples (Mazure, 1998) and more negative events than those with schizophrenia and bipolar depression across a wide range of studies (Paykel, 1994). In instances where depression is preceded by negative events, the onset of depression tends to be rapid, often occurring within one month of the event (Kendler et al., 1999 [double check and find others]).

This strong connection between negative life events and depression along with the need and distress that likely accompanies these events has led to the suggestion that this relationship is causal (Kendler et al., 1999). Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, which shows that the connection is unlikely to be driven solely by depressed individuals creating situations where negative events are likely to be common (Hammen, 2005; Kendler et al., 1999). In addition to this, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict (Kendler & Baker, 2007; Kendler et al., 1999).

@@

Suicide attempts also tend to co-occur with social adversity (Hagen et al., 2008 + cite and describe), with evidence of this existing cross-culturally. Grief over the loss of a loved one (Firth; Devries et al. 2011), bullying (Alavi et al., 2017; Butler et al., 2019; Ford et al., 2017; Wang et al., 2020), and instances of social transgression (Firth; Malinowski; Syme and Hagen 2018; find Western citation to confirm) have all been reported to lead to suicidal behavior in both industrialized (confirm the transgression part) and traditional societies. Of the recorded instances of suicide attempts in the Human Relations Area Files (HRAF), 77.4% of attempts were found to be linked to times of social conflict, 88.7% were linked to fitness threats in general, and 84.9% were linked to situations in which attempters were powerless to resolve their situation alone (Syme et al., 2016). As with depression, assault is an particularly powerful risk factor for suicidal behavior based on multiple large-scale multi-country studies (Devries et al. 2011; Ellsberg et al. 2008) and more localized examinations. For example, both sexual and non-sexual assault were also associated with more self-harm and suicidal behaviors among both indigenous Sami and majority Norwegian adolescents, with those who attempted suicides also reporting more than twice the number of negative life events than those who did not (Reigstad and Kvernmo 2017).

@@

Beyond resulting in the cost differential that allows for costly signaling in the first place, the costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial. Importantly, there are likely few forms of social adversity that do not come with conflict. For example, breakups in both romantic and cooperative relationships are often preceded by conflict (Amato & Hohmann-Marriott, 2007), and surely involve conflict when one side wants reconciliation and the other does not. Likewise, the death of a loved one may also be a source of conflict as the survivor may turn to others for more support who might not want to invest more in the survivor (Hagen, 2003). Regardless of the source of conflict, greater conflict should correspond with greater conflicts of interests, something which would favor greater investment in manipulating others through fake signals. In these instances, less costly signals like verbal requests, facial sadness, and crying are likely to be unreliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals in those experiencing need they cannot resolve on their own.

In addition to signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression and suicide attempts are suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

@@

Beneficial social changes are also reported in traditional societies. Unlike many Western countries, where suicide is often viewed as pathological (Hidaka, 2012), members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness (Shostak, 1981). For example, attempts have been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers (Gutierrez de Pineda & Muirden, 1948; Hilger, 1957; Karsten, 1935; Tessmann, 1930; Wilson, 1960), suggesting that help is at least occasionally provided or reliably expected. In support of this, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, even though suicide was rarely if ever the main topic (Syme et al., 2016).
